---
title: "Concept Overview: Modelling the World with ontologySpecR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concept Overview: Modelling the World with ontologySpecR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ontologySpecR)
```

## What problem does this solve?

Imagine your organisation manages airports, airlines, and flight routes. Data
lives in several databases. Teams build dashboards, run analyses, and write
operational scripts — each with its own idea of what an "airport" is, which
columns matter, and how routes connect to airlines.

The result: **no single source of truth for meaning.** Column names drift,
join logic is duplicated, and a change in one table quietly breaks three
reports.

`ontologySpecR` solves this by providing a **shared vocabulary** — a small
set of building blocks that let you describe *what things exist*, *how they
relate*, and *what can be done to them*. That description is called an
**ontology**, and a complete, versioned snapshot of one is called a
**bundle**.

Think of it like a blueprint for a building. The blueprint does not *contain*
the bricks; it describes their shapes, sizes, and how they fit together. Any
team that reads the blueprint can work from the same understanding.

## The five building blocks

Every ontology in `ontologySpecR` is built from five primitives:

| Building block     | What it describes                        | Real-world analogy            |
|--------------------|------------------------------------------|-------------------------------|
| **Object type**    | A kind of thing (entity)                 | A table's schema              |
| **Link type**      | A named relationship between two things  | A foreign key / join          |
| **Interface**      | A shape contract several types share     | A trait / protocol            |
| **Action type**    | An operation that changes things          | A stored procedure / endpoint |
| **Query**          | A reusable, parameterised question        | A saved report / view         |

We will walk through each one using a small aviation example.

## Object types: describing things

An **object type** says: *"This kind of thing exists, it has these
properties, and this is how you identify one."*

```{r object-type}
airport <- object_type(
  id = "Airport",
  properties = list(
    property_def("airport_id", "string", nullable = FALSE),
    property_def("name",       "string"),
    property_def("city",       "string"),
    property_def("country",    "string"),
    property_def("latitude",   "number"),
    property_def("longitude",  "number")
  ),
  primary_key = "airport_id",
  display_name = "Airport",
  implements = "GeoLocated",
  source_kind = "table",
  source_table = "airports"
)

airport
```

A few things to notice:

- **`property_def()`** defines one column. It needs an `id` and a `type`
  (`"string"`, `"integer"`, `"number"`, `"boolean"`, `"date"`,
  `"datetime"`, or `"json"`). You can mark it as not nullable.
- **`primary_key`** accepts a simple character vector — `"airport_id"` — and
  wraps it automatically. Composite keys work too:
  `primary_key = c("region", "code")`.
- **`source_kind`** and **`source_table`** describe *where the data lives*,
  without coupling you to a specific database. The ontology says "this is a
  table called `airports`"; the runtime decides how to connect to it.

Let's define a second object type so we have something to link:

```{r airline}
airline <- object_type(
  id = "Airline",
  properties = list(
    property_def("airline_id", "string", nullable = FALSE),
    property_def("name",       "string", nullable = FALSE),
    property_def("country",    "string"),
    property_def("active",     "boolean")
  ),
  primary_key = "airline_id",
  display_name = "Airline",
  source_kind = "table",
  source_table = "airlines"
)

airline
```

And a third — a flight route that connects airports to airlines:

```{r route}
route <- object_type(
  id = "FlightRoute",
  properties = list(
    property_def("route_id",       "string",  nullable = FALSE),
    property_def("origin_id",      "string",  nullable = FALSE),
    property_def("destination_id", "string",  nullable = FALSE),
    property_def("airline_id",     "string",  nullable = FALSE),
    property_def("stops",          "integer"),
    property_def("equipment",      "string")
  ),
  primary_key = "route_id",
  display_name = "Flight Route",
  source_kind = "table",
  source_table = "routes"
)

route
```

At this point we have three independent entity definitions. They do not yet
know about each other. That is what links are for.

## Link types: connecting things

A **link type** is a named, typed relationship between two object types. It
says: *"Things of type A are connected to things of type B in this way."*

```{r links}
route_origin <- link_type(
  id = "RouteOrigin",
  from = "FlightRoute",
  to = "Airport",
  cardinality = "many-to-one",
  display_name = "Route Origin",
  join_from_keys = "origin_id",
  join_to_keys = "airport_id"
)

route_destination <- link_type(
  id = "RouteDestination",
  from = "FlightRoute",
  to = "Airport",
  cardinality = "many-to-one",
  display_name = "Route Destination",
  join_from_keys = "destination_id",
  join_to_keys = "airport_id"
)

route_operator <- link_type(
  id = "RouteOperator",
  from = "FlightRoute",
  to = "Airline",
  cardinality = "many-to-one",
  display_name = "Route Operator",
  join_from_keys = "airline_id",
  join_to_keys = "airline_id"
)

route_origin
```

Key ideas:

- **`from`** and **`to`** reference object type IDs — not R objects. This
  keeps the contract declarative and serializable.
- **`cardinality`** can be `"one-to-one"`, `"one-to-many"`, `"many-to-one"`,
  or `"many-to-many"`.
- **`join_from_keys`** / **`join_to_keys`** describe how the link materialises
  at query time. A downstream query engine (like a future `objectSetsR`
  package) reads this to generate the correct SQL join.

With three links defined, we have a small graph:

```
Airport <--[RouteOrigin]-- FlightRoute --[RouteDestination]--> Airport
                                 |
                          [RouteOperator]
                                 |
                                 v
                              Airline
```

## Interfaces: polymorphism across types

What if several object types share a common shape? Airports have
`latitude`/`longitude`. So might weather stations, warehouses, or delivery
depots. Rather than writing separate logic for each, we define an
**interface** — a contract that says *"any type implementing me must have
these properties."*

```{r interface}
geo_located <- interface_type(
  id = "GeoLocated",
  display_name = "Geo-Located",
  display_description = "Any entity with geographic coordinates.",
  required_properties = list(
    property_requirement("latitude",  "number"),
    property_requirement("longitude", "number")
  )
)

geo_located
```

Our `Airport` type already declared `implements = "GeoLocated"`. In a
runtime system, you could ask: *"Give me everything that is GeoLocated"* and
get back airports, weather stations, and warehouses — handled uniformly.

Interfaces can also require links and actions:

```{r interface-complex}
# A more complex interface: anything that can be deactivated
deactivatable <- interface_type(
  id = "Deactivatable",
  display_name = "Deactivatable",
  required_properties = list(
    property_requirement("active", "boolean")
  ),
  required_actions = "Deactivate"
)

deactivatable
```

This says: *"If a type claims to be Deactivatable, it must have an `active`
property and support a `Deactivate` action."* Interface conformance can be
checked programmatically, catching mismatches early.

## Action types: doing things

So far the ontology is static — it describes the world but does not change
it. **Action types** define operations that modify objects, with declared
parameters, effects, and optionally an implementation pointer.

```{r action}
update_status <- action_type(
  id = "UpdateAirportStatus",
  targets = "Airport",
  display_name = "Update Airport Status",
  parameters = list(
    parameter_def("new_status", "string", required = TRUE,
                  display_name = "New Status")
  ),
  effects = list(
    effect_def("update", object_type_id = "Airport",
               notes = "Sets the operational status field.")
  ),
  impl_kind = "r",
  impl_entrypoint = "update_airport_status"
)

update_status
```

Breaking this down:

- **`targets`** lists which object types this action applies to.
- **`parameters`** define the inputs a caller must provide.
- **`effects`** are a high-level declaration of *what changes*: `"create"`,
  `"update"`, `"delete"`, or `"emit"` (for events). This lets governance
  tools reason about what an action can do without reading its implementation.
- **`implementation`** points to the actual code. `kind = "r"` and
  `entrypoint = "update_airport_status"` means: *"call the R function
  `update_airport_status()`."* Other backends (`"sql"`, `"http"`,
  `"plugin"`) are equally valid.

Actions are the bridge between the ontology (descriptive) and operations
(imperative). They are what turn a data model into something you can safely
act on.

## Queries: asking reusable questions

A **query** is a saved, parameterised question that returns object sets,
tables, or scalars. Where actions *change* the world, queries *read* it.

```{r query}
nearby <- query_def(
  id = "NearbyAirports",
  returns_kind = "objectSet",
  returns_object_type_id = "Airport",
  display_name = "Nearby Airports",
  parameters = list(
    parameter_def("lat",       "number", required = TRUE),
    parameter_def("lon",       "number", required = TRUE),
    parameter_def("radius_km", "number", required = TRUE)
  ),
  def_kind = "sql",
  def_body = paste(
    "SELECT * FROM airports",
    "WHERE haversine(latitude, longitude, :lat, :lon) < :radius_km"
  )
)

nearby
```

Queries share the same `parameter_def()` building blocks as actions, so
tooling that presents parameter forms can reuse the same UI code.

## Bundles: packaging it all together

A **bundle** wraps every building block into a single, versioned,
distributable unit. Think of it as a snapshot of your ontology at a point in
time.

```{r bundle}
aviation <- bundle(
  bundle_id = "aviation-demo",
  bundle_version = "0.1.0",
  metadata = list(
    name = "Aviation Demo Ontology",
    description = "Airports, airlines, and flight routes.",
    authors = list("Cathal Byrne"),
    tags = list("aviation", "demo", "transport")
  ),
  objects    = list(airport, airline, route),
  links      = list(route_origin, route_destination, route_operator),
  interfaces = list(geo_located, deactivatable),
  actions    = list(update_status),
  queries    = list(nearby)
)

aviation
```

The bundle's `specVersion` tracks which version of the ontologySpecR schema
it conforms to, while `bundleVersion` tracks your own ontology's evolution.

## Serialization: JSON in, JSON out

Bundles serialize to JSON so they can be stored, versioned in Git, shared
across teams, or served from an API.

```{r serialize}
# Convert to a JSON string
json <- bundle_to_json(aviation, pretty = TRUE)
cat(substr(json, 1, 600), "...\n")
```

And back again:

```{r deserialize}
# Round-trip: JSON -> bundle -> print
restored <- bundle_from_json(json)
restored
```

The round-trip is lossless. Every S3 object is reconstructed with its
correct class, so `is_object_type(restored$objects[[1]])` returns `TRUE`.

For file I/O, use `write_bundle()` and `read_bundle()`:
```r
write_bundle(aviation, "aviation-demo.json")
b <- read_bundle("aviation-demo.json")
```

## Validation: checking the contract

Every bundle can be validated against the built-in JSON Schema (JSON Schema
Draft 2020-12). This catches structural errors — missing required fields,
wrong types, invalid IDs — before they cause problems downstream.

```{r validate}
# Returns TRUE if the bundle is valid
validate_bundle(aviation)
```

If something is wrong, you get actionable error messages instead:

```{r validate-bad}
# A bundle missing required fields
bad_json <- '{"bundleId": "oops"}'
errors <- validate_bundle(bad_json)
errors
```

Validation is especially useful in CI pipelines: *"does this ontology file
still conform to the spec before we deploy?"*

## How the pieces fit together

Here is a visual summary of the five building blocks and their
relationships:

```
  +-----------+     implements     +-------------+
  | Interface |<-------------------| Object Type |
  +-----------+                    +------+------+
       |                                  |
  requires actions                   properties
       |                            primary key
       v                            source binding
  +-------------+                         |
  | Action Type |---- targets ----------->+
  +------+------+                         |
         |                                |
    parameters                     +------+------+
    effects                        |  Link Type  |
    implementation                 +-------------+
                                     from -> to
  +----------+                      cardinality
  |  Query   |                      join keys
  +----------+
    parameters
    returns (objectSet/table/scalar)
    definition (sql/r/dsl)

  +---------+
  | Bundle  |  wraps all of the above into one versioned package
  +---------+
```

## Extensions: making it your own

Every building block has an optional `extensions` field — a free-form named
list where you can attach domain-specific metadata without modifying the
core schema.

```{r extensions}
# Attach custom metadata to an object type
airport_ext <- object_type(
  id = "Airport",
  properties = list(
    property_def("airport_id", "string", nullable = FALSE),
    property_def("name", "string")
  ),
  primary_key = "airport_id",
  extensions = list(
    governance = list(owner = "data-team", review_cycle = "quarterly"),
    ui = list(default_sort = "name", row_limit = 5000L)
  )
)

# Extensions survive serialization
airport_ext$extensions$governance$owner
```

This is how downstream packages can add their own concerns (governance
policies, UI hints, audit trails) while keeping the core spec stable.

## The source mapping model

Object types and properties can declare *where their data comes from*
without specifying *how to connect*. This separation is deliberate.

At the **object type** level:

```r
object_type(
  id = "Airport",
  # ...
  source_kind = "table",    # Could be "table", "view", "query", or "external"
  source_table = "airports",
  source_schema = "public"
)
```

At the **property** level:

```r
property_def(
  "elevation_metres", "number",
  source_column = "elevation_ft",
  source_expression = "elevation_ft * 0.3048"
)
```

This says: *"The ontology property `elevation_metres` comes from the backend
column `elevation_ft`, transformed by multiplying by 0.3048."* A downstream
query engine uses this mapping to generate the correct SQL or dplyr
expression.

The ontology describes the *logical* model. The *physical* connection (which
database, which credentials) is a runtime concern handled by whichever
execution package you choose.

## Where ontologySpecR fits in the ecosystem

`ontologySpecR` is the foundation — the shared contract. It does not query
databases, build UIs, or enforce security. Those are the jobs of
downstream packages that speak the same language:

```
                         ontologySpecR
                     (type system + JSON schema)
                              |
          +-------------------+-------------------+
          |                   |                   |
     ontologyR          objectSetsR          vertexR
  (governance,       (lazy queries,       (system graphs,
   audits, drift)     traversal)           simulation)
          |                   |                   |
          +-------------------+-------------------+
                              |
                   objectExplorerR, workshopR
                     (interactive UIs)
```

Each package depends on `ontologySpecR` for the shared data structures —
and on nothing else from the ecosystem. You can use `ontologySpecR` +
`objectSetsR` without ever installing `vertexR`, or use it standalone to
define and validate ontology bundles.

## Concept definitions: versioned, scoped SQL expressions

A **concept definition** (`concept_def`) is a versioned, scoped SQL
expression that evaluates a boolean concept against an object type.
It is the portable representation consumed by `conceptR` downstream.

```{r concept_def}
busy_airport <- concept_def(
  id             = "busy_airport",
  object_type_id = "Airport",
  scope          = "operations",
  version        = 1L,
  sql_expr       = "daily_passengers > 50000",
  status         = "draft",
  display_name   = "Busy Airport",
  display_description = "An airport handling more than 50,000 passengers per day."
)

busy_airport
```

Key fields:

- **`id`** — a stable identifier; version this upward, never change the id.
- **`object_type_id`** — which object type the SQL expression is written
  against.
- **`scope`** — a context bucket (`"clinical"`, `"operations"`, …) that lets
  multiple teams define concepts for the same object type without collision.
- **`version`** — a monotonically increasing integer within `(id, scope)`.
  Activating a new version is the mechanism for rolling out definition
  changes.
- **`sql_expr`** — the SQL expression. It must return a boolean and is
  evaluated against the object type's backing table by `conceptR`.
- **`status`** — lifecycle state: `"draft"`, `"active"`, or `"deprecated"`.
  Governance transitions are managed by `auditR`.

You can also trace a definition back to an external standard:

```{r concept_def_standard}
concept_def(
  id              = "minimum_runway_length",
  object_type_id  = "Airport",
  scope           = "regulatory",
  version         = 1L,
  sql_expr        = "longest_runway_m >= 1800",
  source_standard = "ICAO"
)
```

### Adding concept definitions to a bundle

Concept definitions are first-class citizens of the bundle:

```{r bundle_with_concepts}
aviation_extended <- bundle(
  bundle_id = "aviation-demo",
  bundle_version = "0.2.0",
  objects    = list(airport, airline, route),
  links      = list(route_origin, route_destination, route_operator),
  interfaces = list(geo_located),
  actions    = list(update_status),
  queries    = list(nearby),
  concepts   = list(busy_airport),
  templates  = list()
)

aviation_extended
```

### Serialisation and deserialisation

Concept definitions serialise and deserialise exactly like every other
bundle component:

```{r concept_roundtrip}
json2 <- bundle_to_json(aviation_extended, pretty = TRUE)
cat(substr(json2, nchar(json2) - 600, nchar(json2)), "\n")
```

```{r concept_deserialise}
restored2 <- bundle_from_json(json2)

# The concept comes back as an ontology_concept_def
is_concept_def(restored2$concepts[[1]])
restored2$concepts[[1]]$id
restored2$concepts[[1]]$scope
```

### Concept templates

A **concept template** (`concept_template_def`) defines a parameterised
SQL expression using `{{param_name}}` placeholders. Downstream consumers
(e.g. `conceptR`) substitute parameter values to instantiate concrete
`concept_def` objects at runtime.

```{r concept_template}
utilisation_template <- concept_template_def(
  id             = "utilisation_threshold",
  object_type_id = "Airport",
  base_sql_expr  = "utilisation_rate > {{threshold}}",
  parameters     = list(
    parameter_def("threshold", "number", required = FALSE,
                  display_name = "Threshold",
                  display_description = "Utilisation rate threshold (default 0.8)")
  ),
  display_name   = "Utilisation Threshold"
)

utilisation_template
```

A `concept_def` can reference a template by setting `template_id` and
supplying `parameter_values`. This records the provenance of derived
definitions:

```{r concept_from_template}
high_utilisation <- concept_def(
  id               = "high_utilisation_airport",
  object_type_id   = "Airport",
  scope            = "operations",
  version          = 1L,
  sql_expr         = "utilisation_rate > 0.85",
  template_id      = "utilisation_threshold",
  parameter_values = list(threshold = 0.85),
  display_name     = "High Utilisation Airport"
)

high_utilisation$templateId
high_utilisation$parameterValues
```

### How downstream packages consume concepts

`conceptR` reads the bundle's `concepts` list and compiles each
`concept_def` into a lazy SQL query:

```r
# Pseudocode — requires conceptR installed
cpt_ctx <- conceptR::concept_context(bundle, con)
result  <- conceptR::cpt_evaluate(cpt_ctx, concept_id = "busy_airport")
# result is a lazy ObjectSet — no rows pulled until materialised
```

`auditR` then controls the lifecycle: sampling disagreements, detecting
drift, and gating activation of new versions behind evidence thresholds.

---

## Summary

| Concept           | Constructor                 | Purpose                                         |
|-------------------|-----------------------------|-------------------------------------------------|
| Property          | `property_def()`            | One typed field on an object type                |
| Primary key       | `primary_key_def()`         | Which properties uniquely identify an object     |
| Object type       | `object_type()`             | Schema for a real-world entity                   |
| Link type         | `link_type()`               | Named relationship between object types          |
| Interface         | `interface_type()`          | Shape contract that multiple types can implement |
| Action type       | `action_type()`             | Parameterised operation with declared effects    |
| Query             | `query_def()`               | Saved, parameterised question                    |
| Concept def       | `concept_def()`             | Versioned, scoped SQL boolean for an object type |
| Concept template  | `concept_template_def()`    | Parameterised SQL template for concept families  |
| Bundle            | `bundle()`                  | Versioned package of all the above               |
| Serialization     | `bundle_to_json()`, `read_bundle()`, etc. | JSON interchange      |
| Validation        | `validate_bundle()`         | Check a bundle against the JSON Schema           |

The core idea is simple: **describe your domain once, in code, and let every
tool and team work from the same specification.**
